---
title: "Miniproject01"
subtitle: "Gourmet Cheeseburgers Across the Globe"
author: "Racheal Alfred"
date: today
format:
  html:
    embed-resources: true
    code-fold: true
    code-summary: "Show Code"
    toc: true
    toc-depth: 3
execute:
  warning: false
  message: false
---

# Executive Summary

This report analyzes Netflix's Top 10 data to highlight the streaming giant's most successful programming across the globe. Our analysis reveals compelling success stories, from the cultural phenomenon of *Squid Game* to Netflix's remarkable growth in the Indian market.

# Introduction

As Netflix continues its mission to produce “gourmet cheeseburgers” — high-quality content with mass appeal — understanding viewership patterns becomes crucial.

# Data Acquisition
```{r setup}
# Create directory structure
if(!dir.exists(file.path("data", "mp01"))){
  dir.create(file.path("data", "mp01"), showWarnings=FALSE, recursive=TRUE)
}

# Define file paths
GLOBAL_TOP_10_FILENAME <- file.path("data", "mp01", "global_top10_alltime.csv")
COUNTRY_TOP_10_FILENAME <- file.path("data", "mp01", "country_top10_alltime.csv")

# Download global and country data if missing
if(!file.exists(GLOBAL_TOP_10_FILENAME)){
  download.file("https://www.netflix.com/tudum/top10/data/all-weeks-global.tsv",
                destfile=GLOBAL_TOP_10_FILENAME)
}

if(!file.exists(COUNTRY_TOP_10_FILENAME)){
  download.file("https://www.netflix.com/tudum/top10/data/all-weeks-countries.tsv",
                destfile=COUNTRY_TOP_10_FILENAME)
}
```

# Data Import and Preparation
```{r libraries}
# Load required libraries
if(!require("tidyverse")) install.packages("tidyverse")
if(!require("DT")) install.packages("DT")
if(!require("lubridate")) install.packages("lubridate")

library(readr)
library(dplyr)
library(stringr)
library(DT)
library(lubridate)
```

## Task 01: Import Global Data
```{r import-global}
GLOBAL_TOP_10 <- read_tsv(GLOBAL_TOP_10_FILENAME)
GLOBAL_TOP_10 <- GLOBAL_TOP_10 |> mutate(season_title = ifelse(season_title == "N/A", NA, season_title))
glimpse(GLOBAL_TOP_10)
```

## Task 02: Import Country Data
```{r import-country}
COUNTRY_TOP_10 <- read_tsv(COUNTRY_TOP_10_FILENAME, na = c("", "NA", "N/A"))
glimpse(COUNTRY_TOP_10)
```

# Initial Data Exploration
```{r helper-functions}
format_titles <- function(df){
  colnames(df) <- str_replace_all(colnames(df), "_", " ") |> str_to_title()
  df
}

GLOBAL_TOP_10 |>
  mutate(`Runtime (Minutes)` = round(60 * runtime)) |>
  select(-season_title, -runtime) |>
  filter(category == "Films (English)") |>
  head(20) |>
  format_titles() |>
  datatable(options=list(searching=FALSE, info=FALSE, pageLength=10)) |>
  formatRound(c("Weekly Hours Viewed", "Weekly Views"), digits=0)
```

# Exploratory Data Analysis

### Task 03: Number of Countries
```{r task-03}
num_countries <- COUNTRY_TOP_10 |> summarise(n_countries = n_distinct(country_name)) |> pull(n_countries)
```
Netflix operates in **`r num_countries`** countries.

### Task 04: Most Popular Non-English Film
```{r task-04}
longest_non_english <- GLOBAL_TOP_10 |>
  filter(category == "Films (Non-English)") |>
  group_by(show_title) |>
  summarise(total_weeks = max(cumulative_weeks_in_top_10)) |>
  arrange(desc(total_weeks)) |>
  slice(1)

film_name <- longest_non_english$show_title
weeks_count <- longest_non_english$total_weeks
```
The non-English film **`r film_name`** spent **`r weeks_count`** weeks in the Top 10.

### Task 05: Longest Film
```{r task-05}
longest_film <- GLOBAL_TOP_10 |> filter(!is.na(runtime)) |> arrange(desc(runtime)) |> slice(1)
longest_film_name <- longest_film$show_title
longest_film_minutes <- round(longest_film$runtime * 60)
```
The longest film in the Top 10 is **`r longest_film_name`**, **`r longest_film_minutes`** minutes long.

### Task 06: Most Viewed Programs by Category
```{r task-06}
most_viewed_by_category <- GLOBAL_TOP_10 |>
  group_by(category, show_title) |>
  summarise(total_hours = sum(weekly_hours_viewed, na.rm=TRUE), .groups="drop") |>
  group_by(category) |> slice_max(total_hours, n=1) |> arrange(category)

most_viewed_by_category |>
  mutate(total_hours = round(total_hours)) |>
  format_titles() |>
  datatable(options=list(searching=FALSE, info=FALSE, pageLength=10)) |>
  formatRound("Total Hours", digits=0)
```

### Task 07: Longest Country Top 10 Run
```{r task-07}
longest_country_run <- COUNTRY_TOP_10 |>
  filter(grepl("TV", category)) |>
  group_by(country_name, show_title, season_title) |>
  summarise(weeks_in_chart = n_distinct(week), .groups="drop") |>
  arrange(desc(weeks_in_chart)) |> head(1)

run_show <- longest_country_run$show_title[1]
run_country <- longest_country_run$country_name[1]
run_weeks <- longest_country_run$weeks_in_chart[1]
if(!is.na(longest_country_run$season_title[1])) run_show <- paste(run_show, "-", longest_country_run$season_title[1])
```
The TV show **`r run_show`** stayed **`r run_weeks`** weeks in **`r run_country`**’s Top 10.

### Task 08: Country with Limited History
```{r task-08}
country_weeks <- COUNTRY_TOP_10 |> group_by(country_name) |> summarise(weeks_of_data = n_distinct(week))
limited_country <- country_weeks |> slice_min(weeks_of_data, n=1)
limited_country_name <- limited_country$country_name
weeks_available <- limited_country$weeks_of_data
last_week <- COUNTRY_TOP_10 |> filter(country_name == limited_country_name) |> summarise(last_date = max(week)) |> pull(last_date)
```
**`r limited_country_name`** has only **`r weeks_available`** weeks of data (ending **`r format(last_week,"%B %Y")`**).

### Task 09: Squid Game Total Viewership
```{r task-09}
squid_game_total <- GLOBAL_TOP_10 |> filter(str_detect(show_title,"Squid Game")) |>
  summarise(total_hours=sum(weekly_hours_viewed,na.rm=TRUE)) |> pull(total_hours)
squid_game_billion <- round(squid_game_total/1e9,2)
```
*Squid Game* logged **`r format(squid_game_total,big.mark=",",scientific=FALSE)`** hours (**`r squid_game_billion`** billion).

### Task 10: Red Notice Views in 2021
```{r task-10}
red_notice_data <- GLOBAL_TOP_10 |> filter(show_title=="Red Notice", year(week)==2021)
if(nrow(red_notice_data)>0){
  total_hours <- sum(red_notice_data$weekly_hours_viewed,na.rm=TRUE)
  runtime_hours <- 1.97
  red_notice_views <- round(total_hours/runtime_hours)
  red_notice_million <- round(red_notice_views/1e6,1)
}else{red_notice_views<-0;red_notice_million<-0}
```
*Red Notice* earned **`r format(red_notice_views,big.mark=",",scientific=FALSE)`** views (**`r red_notice_million`** million).

### Task 11: Films That Climbed to #1 in US
```{r task-11}
us_films_first <- COUNTRY_TOP_10 |> filter(country_iso2=="US", grepl("Film",category)) |>
  group_by(show_title) |> arrange(week) |> slice(1) |> select(show_title,debut_week=week,debut_rank=weekly_rank)
films_that_hit_one <- COUNTRY_TOP_10 |> filter(country_iso2=="US",grepl("Film",category),weekly_rank==1) |> distinct(show_title)
climbers <- inner_join(us_films_first,films_that_hit_one,by="show_title") |> filter(debut_rank>1) |> arrange(desc(debut_week))
num_climbers <- nrow(climbers)
if(num_climbers>0){most_recent_climber<-climbers$show_title[1];most_recent_date<-climbers$debut_week[1];most_recent_debut_rank<-climbers$debut_rank[1]}else{most_recent_climber<-"None";most_recent_date<-Sys.Date();most_recent_debut_rank<-0}
```
**`r num_climbers`** films climbed to #1; latest was **`r most_recent_climber`** (debut #**`r most_recent_debut_rank`**, **`r format(most_recent_date,"%B %Y")`**).

### Task 12: Most International Debut
```{r task-12}
tv_shows <- COUNTRY_TOP_10 |> filter(grepl("TV",category))
show_first_weeks <- tv_shows |> group_by(show_title,season_title) |> summarise(first_global_week=min(week),.groups="drop")
debut_reach <- inner_join(tv_shows,show_first_weeks,by=c("show_title","season_title")) |>
  filter(week==first_global_week) |> group_by(show_title,season_title,first_global_week) |>
  summarise(countries_debuted=n_distinct(country_name),.groups="drop") |> arrange(desc(countries_debuted)) |> head(1)
debut_show <- debut_reach$show_title[1]; if(!is.na(debut_reach$season_title[1])) debut_show<-paste(debut_show,"-",debut_reach$season_title[1])
debut_country_count <- debut_reach$countries_debuted[1]
```
**`r debut_show`** debuted in **`r debut_country_count`** countries — the widest launch.

# Press Releases
## Press Release 1: Stranger Things Season 5

::: {.callout-note icon=false}
## 🎬 STRANGER THINGS PHENOMENON CONTINUES: Final Season Set to Break Records

*Netflix's Crown Jewel Prepares for Epic Finale as Previous Seasons Dominate Global Charts*

{r pr1-analysis}
stranger_things_stats <- GLOBAL_TOP_10 |>
  filter(str_detect(show_title, "Stranger Things")) |>
  summarise(
    total_hours = sum(weekly_hours_viewed, na.rm = TRUE),
    total_weeks = sum(cumulative_weeks_in_top_10, na.rm = TRUE),
    max_weeks = max(cumulative_weeks_in_top_10, na.rm = TRUE)
  )

stranger_countries <- COUNTRY_TOP_10 |>
  filter(str_detect(show_title, "Stranger Things")) |>
  summarise(countries = n_distinct(country_name))

# Compare to other English TV shows
top_english_tv <- GLOBAL_TOP_10 |>
  filter(category == "TV (English)") |>
  group_by(show_title) |>
  summarise(total_hours = sum(weekly_hours_viewed, na.rm = TRUE)) |>
  arrange(desc(total_hours)) |>
  slice(1:5)

st_ranking <- which(top_english_tv$show_title == "Stranger Things")


As Netflix prepares to release the fifth and final season of Stranger Things in late 2025, the series has already cemented its place as one of the most successful television programs in streaming history. With over *r format(stranger_things_stats$total_hours, big.mark=",", scientific=FALSE)* hours viewed globally across its first four seasons, Stranger Things has captured audiences in *r stranger_countries$countries* countries and maintained its position in the Top 10 for an astounding *r stranger_things_stats$total_weeks* cumulative weeks.

The show's unprecedented staying power is evident in its sustained popularity, with individual seasons remaining in the global Top 10 for up to *r stranger_things_stats$max_weeks* consecutive weeks. When compared to other premier English-language TV series on Netflix, Stranger Things ranks in the *top r st_ranking* for total viewership, demonstrating its universal appeal and cultural impact. The series has become a true cultural phenomenon, transcending borders and languages to unite viewers worldwide in their love for Hawkins, Indiana and its supernatural mysteries.

With anticipation building for the final season, Netflix expects to see record-breaking viewership as fans tune in to witness the conclusion of the Upside Down saga. Stranger Things has proven that compelling storytelling and authentic characters can create the "gourmet cheeseburger" effect - premium quality entertainment with mass market appeal.
:::

## Press Release 2: Success in India

::: {.callout-note icon=false}
## 🇮🇳 NETFLIX DOMINATES INDIAN MARKET: Hindi Content Drives Explosive Growth

*Local Language Programming Fuels Subscriber Surge in World's Most Populous Nation*

{r pr2-analysis}
# Analyze Hindi content popularity in India
hindi_programs <- COUNTRY_TOP_10 |>
  filter(country_iso2 == "IN", 
         str_detect(category, "Non-English")) |>
  group_by(show_title, category) |>
  summarise(
    weeks_in_top_10 = n(),
    best_rank = min(weekly_rank),
    .groups = "drop"
  ) |>
  arrange(desc(weeks_in_top_10)) |>
  slice(1:3)

# Get top show that was popular in India but not in US
india_shows <- COUNTRY_TOP_10 |>
  filter(country_iso2 == "IN") |>
  distinct(show_title)

us_shows <- COUNTRY_TOP_10 |>
  filter(country_iso2 == "US") |>
  distinct(show_title)

india_exclusive <- COUNTRY_TOP_10 |>
  filter(country_iso2 == "IN") |>
  filter(!show_title %in% us_shows$show_title) |>
  group_by(show_title) |>
  summarise(
    weeks = n(),
    best_rank = min(weekly_rank)
  ) |>
  arrange(desc(weeks)) |>
  slice(1)

# Calculate growth trend by counting weekly appearances
india_trend <- COUNTRY_TOP_10 |>
  filter(country_iso2 == "IN") |>
  mutate(quarter = quarter(week), year = year(week)) |>
  group_by(year, quarter) |>
  summarise(entries_count = n(), .groups = "drop") |>
  arrange(year, quarter)

recent_growth <- round(
  (tail(india_trend$entries_count, 1) / head(india_trend$entries_count, 1) - 1) * 100
)


Netflix's investment in Hindi-language content is paying enormous dividends in India, with the platform maintaining consistent Top 10 presence and growing engagement. Hits like *r hindi_programs$show_title[1]* have remained in India's Top 10 for *r hindi_programs$weeks_in_top_10[1]* weeks, achieving a best rank of *#r hindi_programs$best_rank[1]*, demonstrating the massive appetite for locally-produced content in the world's most populous nation.

The success of India-exclusive content - programming that charts in India but not in Western markets - highlights Netflix's successful localization strategy. *r india_exclusive$show_title, a title that resonated deeply with Indian audiences while remaining off US charts, spent *r india_exclusive$weeks** weeks in India's Top 10, showcasing the unique preferences of the Indian market and validating Netflix's commitment to diverse, regional storytelling.

Chart activity in India has grown by *r recent_growth%* since tracking began, with consistent quarter-over-quarter increases in Top 10 entries. This growth trajectory positions India as one of Netflix's most important markets for the coming decade, with substantial opportunities for further expansion through continued investment in Hindi and other regional language content.
:::

## Press Release 3: Global Non-English Content Dominance

::: {.callout-note icon=false}
## 🌍 NETFLIX BREAKS LANGUAGE BARRIERS: International Content Achieves Global Success

*Non-English Programming Sees Explosive Growth, Reaching Audiences in Nearly 100 Countries*

{r pr3-analysis}
# Analyze non-English content success globally
non_english_global <- GLOBAL_TOP_10 |>
  filter(category %in% c("Films (Non-English)", "TV (Non-English)")) |>
  summarise(
    total_hours = sum(weekly_hours_viewed, na.rm = TRUE),
    unique_titles = n_distinct(show_title)
  )

# Find most successful non-English show in multiple countries
non_english_countries <- COUNTRY_TOP_10 |>
  filter(str_detect(category, "Non-English")) |>
  group_by(show_title) |>
  summarise(
    countries = n_distinct(country_name),
    weeks_charted = n()
  ) |>
  arrange(desc(countries)) |>
  slice(1)

# Compare English vs Non-English content trends
content_comparison <- GLOBAL_TOP_10 |>
  mutate(
    language_type = ifelse(str_detect(category, "English"), "English", "Non-English"),
    year = year(week)
  ) |>
  group_by(year, language_type) |>
  summarise(yearly_hours = sum(weekly_hours_viewed, na.rm = TRUE), .groups = "drop") |>
  arrange(year, language_type)

# Calculate growth in non-English content
non_english_growth <- content_comparison |>
  filter(language_type == "Non-English") |>
  summarise(
    first_year = first(yearly_hours),
    last_year = last(yearly_hours),
    growth = round((last_year / first_year - 1) * 100)
  )


Netflix's bold bet on international content is reshaping global entertainment, with non-English programming now competing head-to-head with traditional Hollywood productions. Non-English titles have accumulated over *r format(non_english_global$total_hours, big.mark=",", scientific=FALSE)* hours of global viewership across *r non_english_global$unique_titles* unique titles, with breakout hits like *r non_english_countries$show_title* charting in an remarkable *r non_english_countries$countries* countries simultaneously.

This cross-cultural success story demonstrates Netflix's unique ability to identify and promote content that transcends language barriers. Shows from Korea, Spain, France, and beyond now regularly appear in Top 10 lists from São Paulo to Stockholm, from Mumbai to Mexico City. Non-English content viewership has grown by *r non_english_growth$growth%* since tracking began, far outpacing the growth of English-language programming and signaling a fundamental shift in global viewing habits.

Netflix's investment in international production and its willingness to promote non-English content globally represents a strategic masterstroke in the streaming wars. By recognizing that compelling storytelling can overcome language barriers - especially with high-quality subtitles and dubbing - Netflix has unlocked content categories that appeal to increasingly diverse, globally-minded viewers. As other services focus primarily on English-language content, Netflix's diverse international portfolio positions it as the truly global entertainment platform, ready to serve audiences wherever they are and whatever language they speak.
:::

# Conclusion

This analysis of Netflix's Top 10 data reveals a streaming service firing on all cylinders across multiple dimensions: cultivating long-running hits like Stranger Things, successfully penetrating growth markets like India with localized content, and identifying cross-cultural phenomena in international programming. These successes validate Netflix's "gourmet cheeseburger" philosophy - creating high-quality content that nevertheless achieves massive mainstream appeal.

The data clearly shows that Netflix's future lies not just in more content, but in smarter content: programming tailored to specific markets while maintaining the production values and storytelling quality that can attract global audiences. As the streaming landscape becomes increasingly competitive, Netflix's early investments in diverse, international programming provide a significant competitive advantage.


